This file is part of MOFTy.

MOFTy is free software: you can redistribute it and/or modify it under the
terms of the GNU General Public License version 3 as published by the Free
Software Foundation.

MOFTy is distributed in the hope that it will be useful, but WITHOUT ANY
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR
A PARTICULAR PURPOSE. See the GNU General Public License for more details.

You should have received a copy of the GNU General Public License along with
MOFTy. If not, see http://www.gnu.org/licenses/

Copyright(C) 2026 Maximilian Neumann

```{r}
library(MOFA2)
```

```{r}
mefisto_file <- "../output/output_gbm_mefisto/mefisto_model_trained_fullGP.hdf5"
mefisto <- load_model(mefisto_file, sort_factors = FALSE, verbose = TRUE, remove_inactive_factors = FALSE)
```

```{r}
mefisto <- interpolate_factors(mefisto, mefisto@covariates[[1]])
Z_interpol <- t(get_interpolated_factors(mefisto, only_mean = TRUE)[[1]]$mean)
write.csv(Z_interpol, file = "../output/output_gbm_mefisto/Z_interpolated_fullGP.csv")
```