This file is part of MOFTy.

MOFTy is free software: you can redistribute it and/or modify it under the
terms of the GNU General Public License version 3 as published by the Free
Software Foundation.

MOFTy is distributed in the hope that it will be useful, but WITHOUT ANY
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR
A PARTICULAR PURPOSE. See the GNU General Public License for more details.

You should have received a copy of the GNU General Public License along with
MOFTy. If not, see http://www.gnu.org/licenses/

Copyright(C) 2026 Maximilian Neumann

# scNMT downstream analysis

First replace in `reticulate::use_condaenv("mofa2_analysis", conda = "/Users/maximilianneumann/.local/bin/micromamba", required = TRUE)` the user-specific path to micromamba with the appropriate path on your system.

```{r}
reticulate::use_condaenv("mofa2_analysis", conda = "/Users/maximilianneumann/.local/bin/micromamba", required = TRUE)
library(data.table)
library(purrr)
library(ggplot2)
library(ggtext)
library(ggpubr)
library(ggrepel)
library(Matrix)
library(Rfast2)
library(tidyr)
library(MOFA2)
library(reticulate)
library(MOFAdata)

mpl  <- import("matplotlib.cm")
cols <- mpl$get_cmap("seismic")(seq(0, 1, length.out = 256))
cols <- rgb(cols[,1], cols[,2], cols[,3])

ggplot_theme_NoAxes <- function() {
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
}
```


# Load data
The scNMT dataset (scnmt_data.txt.gz) and metadata (scnmt_sample_metadata.txt) is available from the MEFISTO tutorials: [here](https://figshare.com/s/242916198fde3353f3e6)
```{r}
data <- fread("../input/input_scnmt/scnmt_data.txt")
head(data,n=3)
print(data[,length(unique(feature)),by="view"])
print(data[, .(n_NA = sum(is.na(value)), n_samples = uniqueN(sample)), by = "view"])
sample_metadata <- fread("../input/input_scnmt/scnmt_sample_metadata.txt")
print(colnames(sample_metadata))
```


# Visualize data
```{r}
axis_start_x <- -6.5
axis_start_y <- -10.5
axis_length <- 3

celltype.colors <- c(
  "Epiblast"="darkorchid4",
  "Mesoderm"="orangered2",
  "Primitive_Streak"="cornflowerblue",
  "Endoderm"="gold2",
  "Ectoderm"="magenta3"
)

p <- ggscatter(sample_metadata, x="UMAP1", y="UMAP2", color="lineage10x_2", size = 0.5) +
  scale_color_manual(
    values = celltype.colors,
    labels = c(
      "Epiblast" = "Epiblast",
      "Mesoderm" = "Mesoderm",
      "Primitive_Streak" = "Primitive streak",
      "Endoderm" = "Endoderm",
      "Ectoderm" = "Ectoderm"
    )
  ) +
  # Draw custom x-axis line with arrow
  annotate(
    "segment",
    x = axis_start_x,
    y = axis_start_y,
    xend = axis_start_x + axis_length,
    yend = axis_start_y,
    arrow = arrow(length = unit(0.1, "cm"), type = "closed"),
    linewidth = 0.4
  ) +
  annotate(
    "segment",
    x = axis_start_x,
    y = axis_start_y,
    xend = axis_start_x,
    yend = axis_start_y + axis_length,
    arrow = arrow(length = unit(0.1, "cm"), type = "closed"),
    linewidth = 0.4
  ) +
  annotate("text", x = axis_start_x + axis_length, y = axis_start_y - 0.9, label = "UMAP-1", hjust = 0.5, size = 6 / .pt) +
  annotate("text", x = axis_start_x - 1, y = axis_start_y + axis_length, label = "UMAP-2", angle = 90, hjust = 0.5, size = 6 / .pt) +
  theme_void() + # Start with a completely blank theme
  theme(
    legend.position = c(0.85, 0.25),
    legend.text = element_text(size = 6, margin = margin(l = 0, unit = "pt")),
    legend.title = element_blank(),
    legend.key.width = unit(0.3, "cm"),
    legend.key.height = unit(0.3, "cm"),
  ) +
  guides(color = guide_legend(override.aes = list(size = 1))) + 
  theme(
    plot.margin = margin(r = 5, l = 0, unit = "mm")
  )
ggsave(
    filename = file.path("../paper/scnmt", "scnmt.pdf"),
  plot = p,
    width = 5.6/2.54,
    height = 5.6/2.54,
    device = "pdf",
    bg = "transparent"
)

```


# Load trained MOFTy models
```{r}
model_file_Z <- "../output/output_scnmt_cfm/runs_combined_factors_10/mofty_model_Z.hdf5"
model_file_Z_cfm <- "../output/output_scnmt_cfm/runs_combined_factors_10/mofty_model_Z_cfm.hdf5"
model_file_Z_non_cfm <- "../output/output_scnmt_cfm/runs_combined_factors_10/mofty_model_Z_non_cfm.hdf5"

mofty <- load_model(model_file_Z, sort_factors = FALSE, verbose = TRUE, remove_inactive_factors = FALSE)
mofty_cfm <- load_model(model_file_Z_cfm, sort_factors = FALSE, verbose = TRUE, remove_inactive_factors = FALSE)
mofty_non_cfm <- load_model(model_file_Z_non_cfm, sort_factors = FALSE, verbose = TRUE, remove_inactive_factors = FALSE)
samples_metadata(mofty) <- sample_metadata
samples_metadata(mofty_cfm) <- sample_metadata
samples_metadata(mofty_non_cfm) <- sample_metadata

models <- list(
  mofty = mofty,
  mofty_cfm = mofty_cfm,
  mofty_non_cfm = mofty_non_cfm
)
```


# Variance explained
```{r}
print(get_variance_explained(mofty)[["r2_per_factor"]])
print(get_variance_explained(mofty_cfm)[["r2_per_factor"]])
print(get_variance_explained(mofty_non_cfm)[["r2_per_factor"]])

# 1. Extract variance explained data for each model
r2_mofty <- get_variance_explained(mofty)[["r2_per_factor"]][[1]]
r2_cfm <- get_variance_explained(mofty_cfm)[["r2_per_factor"]][[1]]
r2_non_cfm <- get_variance_explained(mofty_non_cfm)[["r2_per_factor"]][[1]]

# 2. Combine into a single data.table
r2_list <- list(
  "MOFTy combined" = r2_mofty,
  "MOFTy CFM" = r2_cfm,
  "MOFTy stochastic" = r2_non_cfm
)

combined_r2_dt <- rbindlist(lapply(names(r2_list), function(model_name) {
  as.data.table(r2_list[[model_name]], keep.rownames = "factor")[, model := model_name]
}), use.names = TRUE, fill = TRUE)

# 3. Melt to long format
plot_dt <- melt(combined_r2_dt, id.vars = c("factor", "model"), variable.name = "view", value.name = "value")
plot_dt[, value := value]
plot_dt[, factor := gsub("Factor", "F", factor)]
factor_levels <- paste0("F", 1:length(unique(plot_dt$factor)))
y_levels <- c(rev(factor_levels), "Total")
plot_dt[, factor := factor(factor, levels = y_levels)]

# 4. Set factor levels for ordering
plot_dt[, model := factor(model, levels = c("MOFTy combined", "MOFTy CFM", "MOFTy stochastic"), labels = c("Combined", "CFM", "Non-CFM"))]
plot_dt[, view := factor(view, levels = c("RNA", "motif_met", "motif_acc"), labels = c(
  "RNA expression",
  "Motif methylation",
  "Motif accessibility"
  ))]

# Total variance explained per model/view for top-row annotations
r2_total_mofty <- get_variance_explained(mofty)$r2_total[[1]]
r2_total_cfm <- get_variance_explained(mofty_cfm)$r2_total[[1]]
r2_total_non_cfm <- get_variance_explained(mofty_non_cfm)$r2_total[[1]]

r2_total_list <- list(
  "MOFTy combined" = r2_total_mofty,
  "MOFTy CFM" = r2_total_cfm,
  "MOFTy stochastic" = r2_total_non_cfm
)

total_dt <- rbindlist(lapply(names(r2_total_list), function(model_name) {
  data.table(
    model = model_name,
    view = names(r2_total_list[[model_name]]),
    value = r2_total_list[[model_name]]
  )
}), use.names = TRUE)

total_dt[, model := factor(model, levels = c("MOFTy combined", "MOFTy CFM", "MOFTy stochastic"), labels = c("Combined", "CFM", "Non-CFM"))]
total_dt[, view := factor(view, levels = c("RNA", "motif_met", "motif_acc"), labels = c("RNA expression", "Motif methylation", "Motif accessibility"))]
total_dt[, factor_label := factor("Total", levels = y_levels)]
total_dt[, label := sprintf("%.1f", value)]


# 5. Create the plot
p_combined <- ggplot(plot_dt, aes(x = model, y = factor, fill = value)) +
  geom_tile(color = "black", linewidth = 0.25) +
  geom_text(
    data = total_dt,
    aes(x = model, y = factor_label, label = label),
    inherit.aes = FALSE,
    size = 5 / .pt,
    vjust = 1,
    hhjust = 0.5,
    nudge_y = 0.1
  ) +
  geom_text(
    aes(label = sprintf("%.1f", value)),
    size = 5 / .pt,
    color = "black"
  ) +
  facet_grid(~ view, scales = "free_x", space = "free_x") +
  coord_cartesian(clip = "off") +
  geom_text(
    aes(label = sprintf("%.1f", value), color = value > 5),
    size = 5 / .pt
  ) +
  scale_color_manual(values = c(`FALSE` = "black", `TRUE` = "white"), guide = "none") +
  scale_y_discrete(
    drop = FALSE,
    labels = function(x) ifelse(x %in% c("Total"),"", x),
    expand = expansion(add = c(0, 0))
  ) +
  scale_fill_gradientn(
    colours = c("#808080", "#FFFFFF", "#3F007D"),  # Gray → White → Purple
    values = scales::rescale(c(-5, 0, 10), from = c(-5, 10)),
    limits = c(-5, 10),
    breaks = c(-5, 0, 5, 10),
    oob = scales::squish,
    name = "%",
    labels = scales::number_format(accuracy = 1)
  ) +
  labs(
    x = NULL, 
    y = "", 
    title = ""
  ) +
  theme_void() +
  theme(
    text = element_text(color = "black"),
    axis.text.x = element_text(
      size = 6,
      color = "black",
      margin = margin(t = 1.5, b = 1),
      angle = 90,
      hjust = 1,
      vjust = 0.5
    ),
    axis.text.y = element_text(size = 6, hjust=1, margin = margin(t = 0, r = 0.2, b = 0, l = 0), color = "black"),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.spacing.x = unit(0, "lines"),
    strip.text = element_text(size = 6, color = "black"),
    legend.position = "right",
    legend.box.spacing = unit(4, "pt"),
    strip.text.x = element_text(margin = margin(b = 2, t = 0)),
    legend.text = element_text(size = 5, margin = margin(r = 0, l = 1, unit = "pt"), color = "black"),
    legend.key.height = unit(0.3, 'cm'),
    legend.key.width = unit(0.12, 'cm'),    
    legend.ticks = element_line(linewidth = 0),
    legend.ticks.length = unit(0.0, "mm"),
    legend.title = element_text(size = 5, margin = margin(b = 0.5, l = -0.7, unit = "pt"), color = "black")
  )

p_combined <- p_combined +
  theme(
    plot.margin = margin(t = 2, r = 8, b = 2, l = 2, unit = "pt"),
    legend.margin = margin(t = 0, r = 2, b = 0, l = 0, unit = "pt")
  )
ggsave(
    filename = file.path("../paper/scnmt", "scnmt_variance_explained_combined.pdf"),
    plot = p_combined,
    width = 7/2.54,
    height = 5/2.54,
    device = "pdf",
    # dpi = 1200,
    bg = "transparent"
)
```


# Factor visualization and imputed/reconstructed data visualization
```{r}
factors_list <- c(2,4,6)
set.seed(42)  # Reproducible RNG
# Moran's I setup (run once)
coords <- as.matrix(sample_metadata[, .(UMAP1, UMAP2)])
nx <- 53; ny <- 65
min_count <- 1
neighbors <- "queen"

xr <- range(coords[,1]); yr <- range(coords[,2])
hx <- (xr[2]-xr[1]) / nx; hy <- (yr[2]-yr[1]) / ny
ix <- pmin(pmax(floor((coords[,1] - xr[1]) / hx), 0), nx-1)
iy <- pmin(pmax(floor((coords[,2] - yr[1]) / hy), 0), ny-1)

dt_cells <- data.table(row = seq_len(nrow(coords)), ix = ix, iy = iy)
cells <- dt_cells[, .(n = .N), by = .(ix, iy)][n >= min_count]
cells[, cid := .I]
dt_cells <- merge(dt_cells, cells[, .(ix, iy, cid)], by = c("ix","iy"))
m <- nrow(cells)

off <- if (neighbors == "rook") {
  rbind(c(1,0), c(-1,0), c(0,1), c(0,-1))
} else {
  as.matrix(expand.grid(dx = -1:1, dy = -1:1))[ -5, ]
}
key <- with(cells, paste(ix, iy))
key2cid <- setNames(cells$cid, key)
edges_i <- integer(); edges_j <- integer()
for (r in seq_len(nrow(cells))) {
  ixr <- cells$ix[r]; iyr <- cells$iy[r]; cid_r <- cells$cid[r]
  nbrs <- cbind(ixr + off[,1], iyr + off[,2])
  nn_keys <- paste(nbrs[,1], nbrs[,2])
  ok <- nn_keys %in% names(key2cid)
  if (any(ok)) {
    cids <- key2cid[nn_keys[ok]]
    edges_i <- c(edges_i, rep.int(cid_r, length(cids)))
    edges_j <- c(edges_j, as.integer(cids))
  }
}
G <- sparseMatrix(i = edges_i, j = edges_j, x = 1, dims = c(m, m))
G <- (G + Matrix::t(G)); diag(G) <- 0
rs <- Matrix::rowSums(G); rs[rs == 0] <- 1
G_row_grid <- as.matrix(Diagonal(x = 1/rs) %*% G)

cid_by_sample <- dt_cells[order(row), cid]
agg_to_grid <- function(x) {
  dt <- data.table(cid = cid_by_sample, val = x)
  out <- dt[, .(val = mean(val, na.rm = TRUE)), by = cid][order(cid)]$val
  if (length(out) != m) stop("Aggregation failed."); out
}

# Calculate and store Moran's I for all models and factors
moran_results <- list()
factors_to_calc <- paste0("Factor", factors_list)

for (model_name in c("mofty", "mofty_cfm", "mofty_non_cfm")) {
  current_model <- models[[model_name]]
  Z <- get_factors(current_model, factors = factors_to_calc)[[1]]
  Z <- Z[sample_metadata$sample, , drop = FALSE] # Align with sample_metadata order
  
  for (f in factors_to_calc) {
    factor_values <- as.numeric(Z[, f])
    x_cell <- agg_to_grid(factor_values)
    mi <- Rfast2::moranI(x_cell, G_row_grid, scaled = TRUE, R = 0)[1]
    moran_results[[paste(model_name, f, sep = "_")]] <- mi
  }
}
moran_results_dt <- as.data.table(moran_results, keep.rownames = "key")

back_ground_fill <- "#e6e6e6ff"   

# Titles for each model row
model_titles <- c(
  mofty = "Combined",
  mofty_cfm = "CFM",
  mofty_non_cfm = "Non-CFM"
)

# Create lists to hold plot components
factor_plots_list <- list()
imputed_reconstructed_plot_rows <- list()


# Create lists to hold plot components
factor_plots_list <- list()
lim <- 1.0

make_row_right <- function(plotlist, ncol = 4) {
  p <- unname(plotlist)
  n_missing <- max(0, ncol - length(p))
  p <- c(rep(list(NULL), n_missing), p)
  cowplot::plot_grid(
    plotlist = p,
    ncol = ncol,
    rel_widths = rep(1, ncol)
  )
}

for (model_name in c("mofty", "mofty_cfm", "mofty_non_cfm")) {
  current_model <- models[[model_name]]
  print(current_model)

  factors.dt <- get_factors(current_model, factors = factors_list)[[1]] %>%
    as.data.table(keep.rownames = TRUE) %>%
    setnames("rn", "sample")

  to.plot_base <- sample_metadata[, c("sample", "UMAP1", "UMAP2")] %>%
    merge(factors.dt, by = "sample") %>%
    melt(id.vars = c("UMAP1", "UMAP2", "sample"), variable.name = "factor")

  # --- New logic: Loop through factors to create individual plots with facet strips ---
  factor_plot_list_inner <- list()
  max_abs_vals_mofty <- list()

  for (f_num in factors_list) {
    f_name <- paste0("Factor", f_num)
    factor_values_comb <- get_factors(models[["mofty"]], factors = f_num)[[1]]
    factor_values_cfm <- get_factors(models[["mofty_cfm"]], factors = f_num)[[1]]
    factor_values_non_cfm <- get_factors(models[["mofty_non_cfm"]], factors = f_num)[[1]]    
    factor_values <- c(factor_values_comb, factor_values_cfm, factor_values_non_cfm)
    max_abs_vals_mofty[[f_name]] <- max(abs(factor_values_comb), na.rm = TRUE)
    
  }
  for (f in paste0("Factor", factors_list)) {
    
    to.plot_factor <- to.plot_base[factor == f]
    to.plot_factor[, value := value / max_abs_vals_mofty[[f]]]
    
    # Create the label for the facet strip
    mi_val <- moran_results[[paste(model_name, f, sep = "_")]]
    factor_title <- sprintf("%s (<i>I</i>: %.2f)", f, mi_val)
    to.plot_factor[, facet_label := factor_title]
    to.plot_factor[, value_abs := abs(value)]
    setorder(to.plot_factor, value_abs)
    to.plot_factor[, value_abs := NULL]

    p_factor <- ggscatter(
      to.plot_factor, x = "UMAP1", y = "UMAP2", fill = "value",
      size = 0.7, shape = 21, color = "black", stroke = 0.05, alpha = 1.0
    ) +
      facet_wrap(~facet_label) + # Use facet_wrap to get the strip box
      scale_fill_gradientn(
        colours = cols,
        name = NULL,
        limits = c(-1, 1),
        breaks = c(-1, 0, 1),
        labels = scales::number_format(accuracy = 0.1),
        oob = scales::squish
      ) +
      theme_void() +
      ggplot_theme_NoAxes() +
      theme(
        panel.grid = element_blank(),
        plot.title = element_blank(), # No main title needed
        strip.background = element_rect(fill = back_ground_fill, color = "black", linewidth = 0.5),
        strip.text = ggtext::element_markdown(size = 6, margin = margin(t = 2, b = 2)),
        axis.line = element_line(linewidth = 0.25),
        legend.position = "none",
        plot.margin = margin(t = 1, r = 1, b = 1, l = 1)
      )
      
    factor_plot_list_inner[[f]] <- p_factor
  }
  
  # Arrange the individual factor plots into a single row
  p_row <- make_row_right(factor_plot_list_inner, ncol = 3)
  
  # Store the row without the title
  factor_plots_list[[model_name]] <- p_row
}

# Arrange the final grid of all model rows
final_factor_plot <- cowplot::plot_grid(
  plotlist = factor_plots_list, 
  ncol = 1
)
imputed_reconstructed_plot_rows[["factors"]] <- final_factor_plot
```

```{r}
# === Part 2 & 3: Imputed and Reconstructed Plots ===
views_to_plot <- c("motif_met", "motif_acc")
target_genes <- c("MSGN1", "FOXA2")
imputed_reconstructed_plot_rows_by_gene <- list()

for (target_gene in target_genes) {
  imputed_reconstructed_plot_rows <- list()

  for (view in views_to_plot) {
    # --- Data Preparation ---
    all_features_in_view <- features_names(mofty)[[view]]
    feature_to_plot <- grep(target_gene, all_features_in_view, value = TRUE)[1]

    if (is.na(feature_to_plot)) {
      stop(paste("No feature containing", shQuote(target_gene), "found in view:", view))
    }

    original_data_vec <- get_data(mofty, views = view, features = setNames(list(feature_to_plot), view))[[1]][[1]][1, ]

    reconstructed_data_list <- list()

    for (model_name in c("mofty", "mofty_cfm", "mofty_non_cfm")) {
      current_model <- models[[model_name]]

      Z_mat_full <- get_factors(current_model)[[1]]
      W_df <- get_weights(current_model, views = view, as.data.frame = TRUE)
      W_feat <- W_df[W_df$feature == feature_to_plot, ]
      common_factors <- intersect(colnames(Z_mat_full), W_feat$factor)
      Z_mat <- as.matrix(Z_mat_full[, common_factors, drop = FALSE])
      W_vec <- setNames(W_feat$value, W_feat$factor)[common_factors]
      reconstructed_data_list[[model_name]] <- as.numeric(Z_mat %*% W_vec)
    }

    to.plot_reconstructed_base <- data.table(
      sample = names(original_data_vec),
      original = original_data_vec
    )
    to.plot_reconstructed_base[[model_titles["mofty"]]] <- reconstructed_data_list[["mofty"]]
    to.plot_reconstructed_base[[model_titles["mofty_cfm"]]] <- reconstructed_data_list[["mofty_cfm"]]
    to.plot_reconstructed_base[[model_titles["mofty_non_cfm"]]] <- reconstructed_data_list[["mofty_non_cfm"]]

    to.plot_reconstructed_base <- to.plot_reconstructed_base %>%
      melt(id.vars = "sample", value.name = "value") %>%
      merge(sample_metadata[, c("sample", "UMAP1", "UMAP2")], by = "sample")

    to.plot_reconstructed_base$variable <- factor(
      to.plot_reconstructed_base$variable,
      levels = c("original", model_titles["mofty"], model_titles["mofty_cfm"], model_titles["mofty_non_cfm"])
    )

    reconstructed_plot_list_inner <- list()
    plot_vars <- levels(to.plot_reconstructed_base$variable)

    mofty_combined_data <- to.plot_reconstructed_base[variable == model_titles["mofty"]]
    mofty_cfm_data <- to.plot_reconstructed_base[variable == model_titles["mofty_cfm"]]
    mofty_non_cfm_data <- to.plot_reconstructed_base[variable == model_titles["mofty_non_cfm"]]
    original_data <- to.plot_reconstructed_base[variable == "original"]

    data_values <- c(
      mofty_combined_data$value,
      mofty_cfm_data$value,
      mofty_non_cfm_data$value,
      original_data$value
    )
    lim <- max(abs(data_values), na.rm = TRUE)

    for (p_var in plot_vars) {
      to.plot_single <- to.plot_reconstructed_base[variable == p_var]
      if (p_var == "original") {
        # Draw NA values first so non-NA points sit on top
        to.plot_single <- to.plot_single[order(is.na(value), decreasing = TRUE)]
      }

      if (p_var == "original") {
        n_total <- length(original_data_vec)
        n_nas <- sum(is.na(original_data_vec))
        na_percentage <- (n_nas / n_total) * 100
        plot_title <- sprintf("Original data<br>(%.1f%% NA)", na_percentage)
      } else {
        data_for_moran <- to.plot_single[match(sample_metadata$sample, sample), ]$value
        x_cell <- agg_to_grid(data_for_moran)
        mi_val <- Rfast2::moranI(x_cell, G_row_grid, scaled = TRUE, R = 0)[1]
        reconstructed_values <- to.plot_single[match(names(original_data_vec), sample), ]$value
        r_squared <- 100 * cor(original_data_vec, reconstructed_values, use = "pairwise.complete.obs")^2
        plot_title <- sprintf(
          "%s<br>(*R*<sup>2</sup>: %.1f%%, <i>I</i>: %.2f)",
          p_var, r_squared, mi_val
        )
      }

      to.plot_single[, facet_label := plot_title]
      to.plot_single[, value_abs := abs(value)]
      setorder(to.plot_single, value_abs)
      to.plot_single[, value_abs := NULL]

      p_single_rec <- ggscatter(
        to.plot_single, x = "UMAP1", y = "UMAP2", fill = "value",
        size = 0.7, shape = 21, color = "black", stroke = 0.05, alpha = 1.0
      ) +
        facet_wrap(~facet_label) +
        scale_fill_gradientn(
          colours = cols,
          name = NULL,
          limits = c(-lim, lim), breaks = c(-lim, 0, lim),
          labels = scales::number_format(accuracy = 0.1),
          oob = scales::squish,
          na.value = "#767676ff"
        ) +
        theme_void() +
        ggplot_theme_NoAxes() +
        theme(
          panel.grid = element_blank(),
          plot.title = element_blank(), # No main title needed
          strip.background = element_rect(fill = back_ground_fill, color = "black", linewidth = 0.5),
          strip.text = ggtext::element_markdown(size = 6, margin = margin(t = 2, b = 2), lineheight = 1.2),
          axis.line = element_line(linewidth = 0.25),
          legend.position = "none",
          plot.margin = margin(t = 1, r = 1, b = 1, l = 1)
        )

      reconstructed_plot_list_inner[[p_var]] <- p_single_rec
    }

    # Arrange the individual plots into a single row
    p_rec_row <- make_row_right(reconstructed_plot_list_inner, ncol = 4)

    # Store the row without the title
    imputed_reconstructed_plot_rows[[paste0(view, "_reconstructed")]] <- p_rec_row
  }

  imputed_reconstructed_plot_rows_by_gene[[target_gene]] <- imputed_reconstructed_plot_rows
}
```

```{r}
# === Create a shared horizontal color bar ===
p_dummy_legend <- ggplot(data.frame(value = c(-1, 1)), aes(x = 1, y = 1, fill = value)) +
  geom_tile() +
  scale_fill_gradientn(
    colours = cols,
    name = NULL, limits = c(-1, 1),
    breaks = c(-0.94, 0, 0.94),
    labels = c("\u2212", "\u0030", "\u002B"),
    guide = guide_colorbar(ticks = FALSE, draw.ulim = FALSE, draw.llim = FALSE)
  ) +
  theme(
    legend.ticks = element_blank(),
    legend.ticks.length = unit(0, "pt")
  ) +
  theme(
    legend.position = c(0.996, 0.3),
    legend.justification = c(1, 0),
    legend.direction = "horizontal",
    legend.key.width = unit(0.3, "cm"), 
    legend.key.height = unit(0.15, "cm"), 
    legend.text = element_text(size = 5, margin = margin(t = 1, r = 2, b = 0, l = 0, unit = "pt")),
    legend.margin = margin(t = 0, r = 0, b = 5, l = 0),
    legend.background = element_rect(fill = "transparent", colour = NA)
  )
shared_colorbar <- cowplot::get_legend(p_dummy_legend)
```


```{r}
factors_plot <- cowplot::plot_grid(
  cowplot::ggdraw() + cowplot::draw_label(model_titles[["mofty"]], hjust = 0.6, vjust = 0.5, fontface = "plain", size = 7),
  factor_plots_list[["mofty"]],
  cowplot::ggdraw() + cowplot::draw_label(model_titles[["mofty_cfm"]], hjust = 0.5, vjust = 0.5, fontface = "plain", size = 7),
  factor_plots_list[["mofty_cfm"]],
  cowplot::ggdraw() + cowplot::draw_label(model_titles[["mofty_non_cfm"]], hjust = 0.5, vjust = 0.5, fontface = "plain", size = 7),
  factor_plots_list[["mofty_non_cfm"]],
  shared_colorbar,
  ncol = 1,
  rel_heights = c(0.1, 1, 0.1, 1, 0.1, 1, 0.2) # Adjusted relative heights
)

# Save the final combined plot
ggsave(
    filename = file.path("../paper/scnmt", "scnmt_factors.pdf"),
    plot = factors_plot,
    width = 10/2.54, # Adjusted width for 4 columns
    height = 15/2.54, # Adjusted height for titles and spacing
    device = "pdf",
    bg = "transparent"
)
```


```{r}
for (target_gene in names(imputed_reconstructed_plot_rows_by_gene)) {
  gene_rows <- imputed_reconstructed_plot_rows_by_gene[[target_gene]]

  imputation_plot <- cowplot::plot_grid(
    cowplot::ggdraw() + cowplot::draw_label(sprintf("Reconstruction and Imputation: %s (methylation)", target_gene), hjust = 0.5, vjust = 0.5, fontface = "plain", size = 7),
    gene_rows[["motif_met_reconstructed"]],
    cowplot::ggdraw() + cowplot::draw_label(sprintf("Reconstruction and Imputation: %s (accessibility)", target_gene), hjust = 0.5, vjust = 0.5, fontface = "plain", size = 7),
    gene_rows[["motif_acc_reconstructed"]],
    shared_colorbar,
    ncol = 1,
    rel_heights = c(0.1, 1, 0.1, 1, 0.25) # Adjusted relative heights
  )

  # Save one imputation figure per target gene
  ggsave(
      filename = file.path("../paper/scnmt", sprintf("scnmt_imputation_%s.pdf", tolower(target_gene))),
      plot = imputation_plot,
      width = 10/2.54, # Adjusted width for 4 columns
      height = 8/2.54, # Adjusted height for titles and spacing
      device = "pdf",
      bg = "transparent"
  )
}
```


# Analyses of weights
```{r}
# Factors to export individually
factor_list <- c(2,6)

w.met <- get_weights(mofty, views="motif_met", factors=factor_list, as.data.frame=TRUE) %>% 
  as.data.table() %>% 
  .[, feature := gsub("_met", "", feature)] %>%
  .[, value := value / max(abs(value)), by = "factor"]

w.acc <- get_weights(mofty, views="motif_acc", factors=factor_list, as.data.frame=TRUE) %>% 
  as.data.table() %>% 
  .[, feature := gsub("_acc", "", feature)] %>%
  .[, value := value / max(abs(value)), by = "factor"]

w.dt <- merge(
  w.met[, c("feature", "factor", "value")], 
  w.acc[, c("feature", "factor", "value")], 
  by = c("feature", "factor")
) %>% .[, feature := strsplit(feature, "_") %>% map_chr(1)]

# Define plot function
factor_plot <- function(factor_id, show_x_axis = TRUE) {
  to.plot <- copy(w.dt[factor == factor_id])
  
  to.label <- copy(w.dt[factor == factor_id]) %>%
    .[, value := abs(value.x) + abs(value.y)] %>%
    setorder(-value) %>% 
    head(n = 10)

  p <- ggplot(to.plot, aes(x = value.x, y = value.y)) +
    geom_point(size = 0.5, color = "#6f99e7", alpha = 1) +
    geom_smooth(method = "lm", color = "#636363", fill = "#afafaf", linewidth = 0.25) +
    coord_cartesian(xlim = c(-1, 1), ylim = c(-1, 1)) +
    scale_x_continuous(breaks = c(-1, 0, 1)) +
    scale_y_continuous(breaks = c(-1, 0, 1)) +
    geom_label_repel(
      data = to.label, aes(label = feature),
      size = 4.5 /.pt, max.overlaps = 100,
      box.padding = 0.07, point.padding = 0.0,
      force = 2, force_pull = 10, direction = "both",
      segment.size = 0.15, segment.color = "black", min.segment.length = 0,
      fill = scales::alpha("#d7bdfa", 0.7),
      linewidth = 0.1,
      label.padding = unit(0.05, "lines"),
      label.r = unit(0.1, "pt")
    ) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.25) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.25) +
    stat_cor(
      aes(label = paste0(
        "italic(r)==", sprintf("%.2f", after_stat(r)), "*','~",
        after_stat(p.label)
      )),
      method = "pearson", label.x = -1, label.y = -1, hjust = 0, size = 5 / .pt
    ) +
    labs(x = "Methylation weights", y = "Accessibility weights") +
    theme_classic() +
    theme(axis.title = element_text(size = 5), axis.text = element_text(size = 5),
          axis.line = element_blank(), axis.ticks = element_line(linewidth = 0.25),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
          plot.margin = margin(2, 2, 2, 2)) +
    annotate("text", x = 1, y = 1, label = factor_id, hjust = 1.1, vjust = 1.1, size = 6 / .pt)
  
  if (!show_x_axis) {
    p <- p + theme(
      axis.title.x = element_text(color = "transparent"),
      axis.text.x  = element_text(color = "transparent"),
      axis.ticks.x = element_line(color = "transparent")
    )
  }
  return(p)
}

# Create and save one plot per factor
for (factor_id in paste0("Factor", factor_list)) {
  p_factor <- factor_plot(factor_id, show_x_axis = TRUE)
  out_name <- sprintf("scnmt_weights_scatter_%s.pdf", tolower(factor_id))
  ggsave(
    filename = file.path("../paper/scnmt", out_name),
    plot = p_factor,
    width = 6.5 / 2.54,
    height = 5.5 / 2.54,
    device = "pdf",
    bg = "transparent"
  )
}
```


## Weight statistics
```{r}
views_for_stats <- c("RNA", "motif_met", "motif_acc")
all_factor_ids <- 1:10

for (view_name in views_for_stats) {
  # Get raw (unscaled) weights for all factors
  w_raw <- get_weights(mofty, views = view_name, factors = all_factor_ids, as.data.frame = TRUE) %>%
    as.data.table()

  stats_list <- lapply(paste0("Factor", all_factor_ids), function(f) {
    vals <- w_raw[factor == f, value]

    # Scaled MAD  (R's mad() uses constant = 1.4826 by default)
    med_val <- median(vals, na.rm = TRUE)
    scaled_mad_val <- mad(vals, na.rm = TRUE)

    # over_mad for every feature: |w - median(w)| / scaled_MAD
    over_mad <- abs(vals - med_val) / scaled_mad_val

    # Top 1% by absolute weight value -> minimum over_mad among them
    n_top <- max(1, ceiling(length(vals) * 0.01))
    top_idx <- order(abs(vals), decreasing = TRUE)[1:n_top]
    min_over_mad_top1 <- min(over_mad[top_idx])

    data.table(
      factor           = f,
      scaled_mad       = round(scaled_mad_val, 6),
      min_over_mad_top1pct = round(min_over_mad_top1, 4)
    )
  })

  stats_dt <- rbindlist(stats_list)

  out_file <- file.path("../paper/scnmt",
                         sprintf("scnmt_%s_weight_stats.csv", tolower(view_name)))
  fwrite(stats_dt, out_file)
  cat(sprintf("\n=== %s ===\nSaved: %s\n", view_name, out_file))
  print(stats_dt)
}
```


# Gene set enrichment analysis (GSEA)
```{r}
data("MSigDB_v6.0_C5_mouse") 
```

```{r echo=FALSE}
features_names(mofty)[["RNA"]] <- toupper(features_names(mofty)[["RNA"]])
```

```{r}
enrichment.out <- run_enrichment(
  object = mofty,
  view = "RNA",
  feature.sets = MSigDB_v6.0_C5_mouse,
  statistical.test = "parametric",
  alpha = 0.05
)
```

```{r}
enrichment.out$set.statistics
```

```{r}
# Save enrichment statistics with p-values (long format)
stats_mat <- enrichment.out$set.statistics
pval_mat <- enrichment.out$pval
padj_mat <- enrichment.out$pval.adj

stats_long <- data.table(
  pathway = rep(rownames(stats_mat), times = ncol(stats_mat)),
  factor = rep(colnames(stats_mat), each = nrow(stats_mat)),
  statistic = as.numeric(stats_mat),
  pval = as.numeric(pval_mat),
  padj = as.numeric(padj_mat)
)
stats_long[, neglog10_padj := -log10(padj)]

stats_csv <- file.path("../paper/scnmt", "scnmt_enrichment_stats_pvals.csv")
fwrite(stats_long, stats_csv)
cat("Saved enrichment statistics with p-values to:", stats_csv, "\n")

factor_list <- c(4)

get_mapped_var <- function(plot_obj, aes_name) {
  extract_var <- function(expr_obj, data_names) {
    if (is.null(expr_obj)) return(NA_character_)
    if (is.symbol(expr_obj)) {
      nm <- as.character(expr_obj)
      if (nm %in% data_names) return(nm)
    }
    vars <- all.vars(expr_obj)
    vars <- vars[vars %in% data_names]
    if (length(vars) > 0) return(vars[1])
    NA_character_
  }

  data_names <- names(plot_obj$data)
  m <- plot_obj$mapping[[aes_name]]
  var_name <- extract_var(m, data_names)
  if (!is.na(var_name)) return(var_name)
  for (ly in plot_obj$layers) {
    ml <- ly$mapping[[aes_name]]
    var_name <- extract_var(ml, data_names)
    if (!is.na(var_name)) return(var_name)
  }
  NA_character_
}

for (i in factor_list) {
  stats_i <- as.numeric(stats_mat[, i])
  names(stats_i) <- rownames(stats_mat)
  sign_map <- ifelse(stats_i > 0, "+", ifelse(stats_i < 0, "-", "0"))
  names(sign_map) <- names(stats_i)

  p <- plot_enrichment(enrichment.out, factor = i, max.pathways = 10)

  x_var <- get_mapped_var(p, "x")
  y_var <- get_mapped_var(p, "y")

  ann <- NULL
  x_is_num <- FALSE
  y_is_num <- FALSE
  x_lim_right <- NA_real_
  y_lim_top <- NA_real_
  if (!is.na(x_var) && !is.na(y_var) && x_var %in% names(p$data) && y_var %in% names(p$data)) {
    x_raw <- p$data[[x_var]]
    y_raw <- p$data[[y_var]]

    x_is_num <- is.numeric(x_raw)
    y_is_num <- is.numeric(y_raw)

    if (x_is_num && !y_is_num) {
      x_vals <- as.numeric(x_raw)
      y_vals <- as.character(y_raw)
      sign_vals <- sign_map[match(y_vals, names(sign_map))]
      sign_vals[is.na(sign_vals)] <- ""

      x_rng <- range(x_vals, na.rm = TRUE)
      x_lim_right <- x_rng[2] 
      ann <- data.frame(x = x_lim_right, y = y_vals, sign_label = sign_vals)
    } else if (y_is_num && !x_is_num) {
      x_vals <- as.character(x_raw)
      y_vals <- as.numeric(y_raw)
      sign_vals <- sign_map[match(x_vals, names(sign_map))]
      sign_vals[is.na(sign_vals)] <- ""

      y_rng <- range(y_vals, na.rm = TRUE)
      y_lim_top <- y_rng[2] + 5
      ann <- data.frame(x = x_vals, y = y_lim_top, sign_label = sign_vals)
    }
  }

  label_wrap <- function(x) {
    x_clean <- gsub("_", " ", x)
    vapply(x_clean, function(label) {
      paste(strwrap(label, width = 40), collapse = "\n")
    }, character(1))
  }

  p <- p +
    labs(y = bquote("Factor" ~ .(i) * ", GSEA (-log adj. " * italic(p) * "-value)")) +
    labs(x = NULL) +
    theme_classic() +
    theme(
      text = element_text(size = 5),
      axis.line = element_line(linewidth = 0.25, color = "black"),
      axis.ticks = element_line(linewidth = 0.25, color = "black"),
      plot.title = element_text(size = 5, hjust = 0.5),
      axis.text.x = element_text(color = "black"),
      axis.text.y = element_text(color = "black"),
      panel.border = element_blank(),
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA)
    )

  if (x_is_num) {
    p <- p +
      scale_y_discrete(labels = label_wrap) +
      scale_x_continuous(limits = c(NA, x_lim_right), expand = expansion(mult = c(0, 0)))
  } else if (y_is_num) {
    p <- p +
      scale_x_discrete(labels = label_wrap) +
      scale_y_continuous(limits = c(NA, y_lim_top), expand = expansion(mult = c(0, 0)))
  } else {
    p <- p + scale_x_discrete(labels = label_wrap)
  }

  if (!is.null(ann)) {
    p <- p + geom_text(
      data = ann,
      aes(x = x, y = y, label = sign_label),
      inherit.aes = FALSE,
      size = 2.0,
      hjust = 1,
      vjust = 0.5,
      color = "black"
    )
  }

  # Make the dot at the end of each bar smaller
  p$layers <- lapply(p$layers, function(l) {
    if (inherits(l$geom, "GeomPoint")) l$aes_params$size <- 0.5
    if (inherits(l$geom, "GeomSegment")) l$aes_params$linewidth <- 0.25
    if (inherits(l$geom, "GeomLinerange")) l$aes_params$linewidth <- 0.25
    l
  })
  p$layers <- p$layers[!vapply(p$layers, function(l) inherits(l$geom, "GeomHline"), logical(1))]

  ggsave(
    filename = file.path("../paper/scnmt", sprintf("scnmt_enrichment_factor%s.pdf", i)),
    plot = p,
    width = 6.7 / 2.54,
    height = 3.8 / 2.54,
    device = "pdf",
    bg = "transparent"
  )
}
```


```{r}
sessionInfo()
```