This file is part of MOFTy.

MOFTy is free software: you can redistribute it and/or modify it under the
terms of the GNU General Public License version 3 as published by the Free
Software Foundation.

MOFTy is distributed in the hope that it will be useful, but WITHOUT ANY
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR
A PARTICULAR PURPOSE. See the GNU General Public License for more details.

You should have received a copy of the GNU General Public License along with
MOFTy. If not, see http://www.gnu.org/licenses/

Copyright(C) 2026 Maximilian Neumann

# scNMT preprocessing

First replace in `reticulate::use_condaenv("mofa2_analysis", conda = "/Users/maximilianneumann/.local/bin/micromamba", required = TRUE)` the user-specific path to micromamba with the appropriate path on your system.

```{r}
reticulate::use_condaenv("mofa2_analysis", conda = "/Users/maximilianneumann/.local/bin/micromamba", required = TRUE)
library(data.table)
library(MOFA2)
```


# Load data
The scNMT dataset (scnmt_data.txt.gz) and metadata (scnmt_sample_metadata.txt) is available from the MEFISTO tutorials: [here](https://figshare.com/s/242916198fde3353f3e6)

```{r}
data <- fread("../input/input_scnmt/scnmt_data.txt")
head(data,n=3)
```

```{r}
sample_metadata <- fread("../input/input_scnmt/scnmt_sample_metadata.txt")
colnames(sample_metadata)
```


# Prepare untrained MEFISTO object
```{r}
outfile <- "../input/input_scnmt/mefisto_model_untrained.hdf5"
mefisto <- create_mofa_from_df(data)
samples_metadata(mefisto) <- sample_metadata %>% 
  .[sample%in%unlist(samples_names(mefisto))] %>%
  setkey(sample) %>% .[unlist(samples_names(mefisto))]
model_opts <- get_default_model_options(mefisto)
mefisto <- set_covariates(mefisto, c("UMAP1","UMAP2"))
train_opts <- get_default_training_options(mefisto) 
train_opts$maxiter <- 0
mefisto_opts <- get_default_mefisto_options(mefisto)
data_opts <- get_default_data_options(mefisto)
data_opts$use_float32 <- FALSE
mefisto <- prepare_mofa(
  mefisto,
  model_options = model_opts,
  mefisto_options = mefisto_opts,
  training_options = train_opts,
  data_options = data_opts
)
mefisto <- run_mofa(mefisto, outfile=outfile)
```

```{r}
sessionInfo()
```
